Cool title here

### 2018-02-11 Frontiers_Manuscript/Villasenor-Derbez.tex
Added english text for intro and methods

\begin{document}
\onecolumn
\firstpage{1}

@@

Nice abstract here

\medskip
\tiny
 \keyFont{ \section{Keywords:} Marine Reserves, Marine Conservation, Small Scale Fisheries, Citizen
Science, Mexico.}

@@

Overfishing and unsustainable fishing practices threaten marine
ecosystems around the world \citep{halpern_2008-dK,halpern_2017-Zi}. The
implementation of marine reserves (\emph{i.e.} areas where all fishing
activities are off--limits; MRs) is a common management approach used to
manage the spatial distribution of fishing effort and recover stocks
\citep{afflerbach_2014-HP,krueck_2017-J1,sala_2017-69}. Recent research
poses that marine reserves can also help buffer climate change
\citep{roberts_2017-J9}, provide a refuge to environmental variability
\citep{micheli_2012-EU}, help reduce by-catch \citep{hastings_2017-sm}
and, in general, increase biomass, richness and densities of organisms
within the protected regions
\citep{lester_2009-Ks,giakoumi_2017-V2,sala_2017-69}. However, the
extent to which MRs yield benefits to fisheries remains relatively
unexplored \citep{krueck_2017-J1} and the focus is often only in the
social and economic dimensions \citep{klein_2008,charles_2008}, rarely
providing a hollistic evaluation of the social-ecological system
\citep{halpern_2013,lpezangarita_2014,mascia_2017-m_}. Here, we combine
causal inference techniques \citep{depalma_2018} and the
social-ecological systems framework
\citep{ostrom_2009-hg,mascia_2017-m_} to provide a comprehensive
evaluation of four community-based marine reserves in three coastal
communities in Mexico.

En México, las reservas marinas han sido comúnmente establecidas como
zonas núcleo dentro de Reservas de la Biósfera (RBs), administradas por
la Comisión Nacional de Áreas Naturales Protegidas (CONANP). Al día de
hoy, 36 RBs protegen una porción del ambiente marino en México. Sin
embargo, solamente 26 de estas incluyen (pequeñas) zonas núcleo donde
las actividades pesqueras están prohibidas. Aunque la CONANP ha hecho
esfuerzos importantes por involucrar a los actores durante la
implementación de las reservas, esto aún se caracteriza por un proceso
descendente, el cual conlleva a la falta de cumplimiento por parte de
los actores. La escasez de recursos monetarios y humanos de la limitan
también el monitoreo y vigilancia de las reservas, y a su vez, el
desempeño de la reserva.

Buscando promover una alternativa con procesos ascendentes para
implementar reservas marinas, las Organizaciones de la Sociedad Civil
(OSCs) comenzaron a trabajar con comunidades pesqueras para establecer
reservas comunitarias \citep{uribe_2010-u2} . Estas son comúnmente
establecidas dentro de zonas de concesión, una forma de derechos de uso
territoriales para pesquerías (TURF, en inglés). Al permitir a los
pescadores diseñar sus propias reservas, una mayor proporción de la
comunidad está de acuerdo con los perímetros y reglas establecidas, y
por lo tanto los respetan
\citep{gelcich_2015-Gw,espinosaromero_2014-PY,beger_2004-Y8} .
Adicionalmente, los pescadores pueden implementar sus reservas por un
periodo acordado (usualmente cinco años), después del cual la reserva
puede ser abierta a la pesca. Esto provee a los pescadores con un
sentido de confianza de que, en caso de ser necesario, aún tienen acceso
a pescar esa zona. Las reservas son directamente vigiladas y
monitoreadas por la comunidad, quienes comúnmente utilizan pequeñas
embarcaciones (\emph{e.g.} pangas) para patrullar la zona, o realizan
avistamientos desde la costa en búsqueda de pescadores ilegales Aún así,
las reservas comunitarias carecen de reconocimiento legal; por lo tanto,
no hay forma de penalizar a los infractores.

Sin embargo, en el 2014 una nueva norma \citep{nom} permite a los
pescadores solicitar el establecimiento de reservas marinas bajo el
nombre de ``Zonas de refugio Pesquero'' (ZRP). El manejo de las ZRP
combina procesos ascendentes y descendentes al reconocer legalmente las
reservas propuestas por las comunidades. Posterior a la revisión por
parte de la Comisión Nacional de Acuacultura y Pesca (CONAPESCA) y la
opinión técnica del Instituto Nacional de Acuacultura y Pesca (INAPESCA)
las ZRP son establecidas por el periodo solicitado por los pescadores.
El monitoreo y la vigilancia de las ZRP es típicamente llevado a cabo
por la comunidad , con ayuda de OSCs locales. Hasta este cambio
regulatorio, las reservas comunitarias no contaban con el soporte legal,
y eran solamente reconocidas por la comunidad. Al día de hoy, existen 39
ZRP establecidas en el Pacífico, Golfo de California y Caribe Mexicano.

Aunque existen tres aproximaciones generales para implementar reservas
marinas en México (\emph{i.e.} Zonas núcleo dentro de AMP, reservas
comunitarias y Zonas de Refugio Pesquero), aún no comprendemos a fondo
las características sociales que permiten su efectividad. La ciencia de
reservas marinas se ha enfocado ampliamente en los efectos biológicos
que estas tienen
\citep{lester_2009-Ks,giakoumi_2017-V2,sala_2017-69,afflerbach_2014-HP,krueck_2017-J1}.
Aunque el aspecto ecológico de las reservas es importante para su éxito,
su efectividad también depende del estado socioeconómico y los sistemas
de gobernanza de las comunidades pesqueras.

La literatura indica que diferentes características influyen en el éxito
de una reserva. En Palau, por ejemplo, la edad (\emph{i.e.} tiempo
transcurrido desde implementación), tamaño y hábitat contenido son
características claves que determinan la efectividad
\citep{friedlander_2017-oI}. Por otro lado, en el Mar Mediterráneo,
\citet{difranco_2016-Xw} identifican que la procuración y vigilancia,
presencia de un plan de manejo, participación de pescadores en el
manejo, representación de pescadores en la toma de decisiones y
promoción de la pesca sustentable son los cinco factores que incrementan
la salud de los stocks y el ingreso económicos a los pescadores, a la
vez que se presenta una mayor aceptación social de las prácticas de
manejo. En una aproximación global, \citet{edgar_2014-UO} encuentran que
la procuración, edad, tamaño y aislamiento son determinantes de la
efectividad de las reservas. Por lo tanto, observamos que las
características que habilitan el éxito varían a través de regiones, y
poco esfuerzo se ha hecho por comprender estas interacciones en México.

El objetivo de este trabajo este trabajo es realizar una evaluación de
la efectividad de reservas marinas en México, presentando resultados de
cinco comunidades costeras como caso de estudio. Con el fin de obtener
una visión holística del sistema, la evaluación se realizará tomando en
cuenta indicadores biológicos, socioeconómicos y de gobernanza. La
evaluación de éstos cinco casos de estudios nos permitirá identificar la
manera en que las características socioeconómicas y de gobernanza se
relacionan con la efectividad (biológica) de las reservas marinas
evaluadas. Los patrones identificados podrán utilizarse para informar la
toma de decisiones para la implementación de la red de reservas marinas
en la Región de las Grandes Islas del Golfo de California.

@@

Las comunidades utilizadas en este reporte se distribuyen a lo largo de
la costa Pacífica de Baja California (n = 1) y el Sistema Arrecifal
Mesoamericano (n = 2; Fig \ref{fig:map}).

\subsubsection{Isla Natividad}\label{isla-natividad}

La Isla Natividad se encuentra en la costa oeste de la Península de Baja
California, donde el hábitat predominante es el bosque de kelp o sargazo
gigante (\emph{Macrocystis pyrifera}) y los arrecifes rocosos. En la
isla, la Sociedad Cooperativa de Producción Pesquera (SCPP) Buzos y
Pescadores de la Baja California SCL realiza actividades de extracción
de los recursos marinos. Aunque la langosta roja (\emph{Panulirus
interruptus}) es la especie más importante en términos económicos, otras
especies importantes incluyen la escama (con un enfoque en Jurel;
\emph{Seriola lalandi}), el pepino de mar (\emph{Parastichopus
parvimensis}), el erizo rojo (\emph{Mesocentrotus franciscanus}), el
caracol (\emph{Megastraea turbanica} y \emph{M. undosa}) y, hasta el
2010, el abulón (\emph{Haliotis sp.}). En 2006, por medio de un proceso
participativo, la cooperativa decidió establecer dos reservas marinas de
manera voluntaria. Agentes externos a la cooperativa, como personal de
Comunidad y Biodiversidad A.C., académicos de la Universidad de
Stanford, y personal de la CONANP (de la oficina de Reserva de la
Biósfera El Vizcaíno), también participaron en el diseño e
implementación de las reservas. Las reservas fueron establecidas como
instrumento de manejo pesquero, buscando recuperar las poblaciones de
abulón y otros invertebrados. Al día de hoy, las reservas marinas de
Isla Natividad no han recibido reconocimiento legal, pero la cooperativa
ha mostrado interés por reconocerlas como Zonas de Refugio Pesquero
(ZRP). Los pescadores tienen un sistema de turnos para vigilar la
reserva día y noche desde embarcaciones patrulla.

\subsubsection{María Elena}\label{maria-elena}

María Elena es una comunidad pesquera en la costa de Quintana Roo. Los
arrecifes coralinos y manglares son los principales ecosistemas
representados en la zona. El campo pesquero es utilizado por pescadores
de la SCPP Cozumel scl (de la Isla de Cozumel). La principal especie
aprovechada por ésta organización es la langosta espinosa del caribe
(\emph{Panulirus argus}). La cooperativa cuenta con permiso de pesca de
escama y concesión de langosta. En el 2012, la Cooperativa, en conjunto
con la Alianza Kanan Kay, COBI, CONANP, CONAPESCA, Oceanus, Fundación
Claudia y Roberto Hernández, Fundación Haciendas del Mundo Maya,
establecieron ocho ZRP con una vigencia de cinco años. La vigilancia de
las reservas se realiza por medio del equipo de vigilancia comunitaria,
con apoyo de la CONANP y una embarcación -donada por COBI- utilizada
para realizar recorridos frecuentes.

\subsubsection{Punta Herrero}\label{punta-herrero}

La comunidad de Punta Herrero se encuentra aproximadamente a 15 km al
sur del campo pesquero de María Elena. De igual manera, los arrecifes
coralinos y manglares son los principales ecosistemas representados en
la zona, y la principal especie explotada es la langosta. Sin embargo,
la SCPP José María Azcorra también cuenta con permisos para pesca de
escama y tiburón y una concesión de langosta. En una réplica del
ejercicio realizado en María Elena -con presencia de los mismos
actores-, cuatro ZRP fueron establecidas en el 2013, con una vigencia de
cinco años. El equipo de vigilancia comunitaria, con apoyo de la CONANP,
se encarga de la vigilancia de las reservas.

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{\label{fig:unnamed-chunk-1}\label{fig:map}Mapa de la localización
general de las comunidades de estudio. El panel de la derecha es un
acercamiento a las comunidades de María Elena y Punta Herrero.}
\end{figure}

@@

Para evaluar las reservas, utilizamos tres fuentes de información. La
información ecológica proviene de los monitoreos ecológicos realizados
anualmente en las zonas reserva y control. Cada año, se realizan censos
visuales para evaluar las comunidades de peces e invertebrados,
registrando riquezas, abundancias y tallas (en peces). Esta información
nos permite calcular los indicadores biológicos de manera anual. Al
tener valores de diferentes indicadores biológicos antes y después de la
implementación de las reservas, para las zonas de reserva y sitios
control, tenemos un diseño muestreal de Antes-Después-Control-Impacto.

También incluimos información socioeconómica relevante proveniente de
los avisos de arribo de CONAPESCA. En este caso, se tienen registros
mensuales de los recursos aprovechados por las diferentes comunidades,
en los que se reportan los arribos (Toneladas) y el valor de los arribos
(\$). La información se encuentra disponible para el periodo 2001 -
2014. Ya que las categorías registradas por CONAPESCA son amplias y
existe un nivel de error, utilizamos únicamente los arribos reportados
para langosta entera fresca a nivel de cooperativa. Los ingresos
generados por arribos son ajustados por medio del índice de precio al
consumidor.

La información de gobernanza fue obtenida a nivel de comunidad, pidiendo
a personas familiares con las comunidades que proveyeran la información
necesaria. La información de gobernanza no es evaluada de manera
cuantitativa. En su lugar, esta se interpreta de manera tal que podamos
comprender qué decisiones, reglas y estructuras tienen un impacto en la
reserva.

@@

Dada la similitud de objetivos entre las reservas, la evaluación se
realiza con los mismos indicadores. En este caso, se utilizan 6
indicadores biológicos, 2 socioeconómicos y 5 de gobernanza (Tabla
\ref{table:indicators}). Según la disponibilidad de datos, se calculó la
densidad de las especies objetivo presentadas en la sección de
descripción de las comunidades. El criterio de selección fue que cada
especie debía tener, por lo menos, dos observaciones anuales para las
zonas de reserva y control.

\begin{table}

\caption{\label{tab:unnamed-chunk-2}\label{table:indicators}Lista de indicadores utilizados para evaluar resvas marinas, agrupados por tipo.}
\centering
\begin{tabular}[t]{l|l}
\hline
Category & Indicador\\
\hline
Biological & Índice de diversidad de shannon\\
\hline
 & Riqueza\\
\hline
 & Densidad\\
\hline
 & Nivel trófico\\
\hline
 & Biomasa\\
\hline
 & Densidad de especies objetivo\\
\hline
Socioeconomic & Ingresos por especies objetivo\\
\hline
 & Arribos de especies objetivo\\
\hline
Governance & Tipo de acceso a la pesquería\\
\hline
 & Grado de pesca ilegal\\
\hline
 & Procuración de la reserva\\
\hline
 & Tipo de organización pesquera\\
\hline
 & Edad de la reserva\\
\hline
\end{tabular}
\end{table}

\subsubsection{Biological}\label{biological}

Utilizando un análisis de diferencia en diferencias podemos estimar el
efecto que la reserva tienen en los indicadores biológicos
\citep{moland_2013-VP} con el uso de un modelo de regresión lineal
múltiple:

\[I = \beta_0 + \sum \gamma Year + \beta_1 Zona + \sum \lambda Year\times Zona + \Omega + \epsilon\]

En este caso, modelamos los años como factores, tomando como referencia
el primer año en la serie de datos de cada comunidad. Modelar los años
como factores reduce la estructura del modelo, y relaja el ajuste al no
asumir una tendencia lineal entre años; es decir, el cambio observado
entre 2006 - 2007 no deberá de ser igual al observado entre el 2009 -
2010. Incluimos también un término para la zona, en la que la variable
toma un valor de 0 si el sitio es una zona control y de 1 si es una zona
de reserva. Finalmente, incluimos un término de interacción entre la
variable de Zona y el Año. En este modelo, \(\lambda_i\) representa el
efecto que la reservas tuvo sobre un indicador en cada año y con
respecto a los sitios control. El termino \(\Omega\) captura efectos
fijos por especies y por sitio.

\subsubsection{Socioeconomic}\label{socioeconomic}

El análisis de datos socioeconómicos se aplicó únicamente a Isla
Natividad, María Elena y Punta Herrero, siguiendo un modelo con la
forma:

\[I = \beta_0 + \beta_1Post\]

Que nos permite comparar el cambio en el promedio de los indicadores
antes (Post = 0) y después (Post = 1) de la implementación de la
reserva. Tanto para los indicadores biológicos como los socioeconómicos,
los coeficientes fueron ajustador con el estimador de muestras
heterocedásticas.

@@

>>>>

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\onecolumn
\firstpage{1}

@@

Abstract here

\medskip
\tiny
 \keyFont{ \section{Keywords:} Marine Reserves, Marine Conservation, Small Scale Fisheries, Citizen
Science, Mexico, Social--Ecological Systems}

@@

Marine ecosystems around the world sustain significant impacts due to
overfishing and unsustainable fishing practices
\citep{halpern_2008-dK,worm_2006-IB,pauly_2005-qV}. A common approach to
manage the spatial distribution of fishing effort and recover stocks is
through the implementation of marine reserves (\emph{i.e.} areas where
all fishing activities are off--limits; MRs)
\citep{afflerbach_2014-HP,krueck_2017-J1,sala_2017-69}.

Marine reserve science has largely focused on understanding the
ecological effects of these areas, which include increased biomass,
richness, and densities of organisms within the protected regions
\citep{lester_2009-Ks,giakoumi_2017-V2,sala_2017-69}, climate change
mitigation \citep{roberts_2017-J9}, and protection from environmental
variability \citep{micheli_2012-EU}. However, there is considerably less
literature focusing on the relationship between socioeconomic and
governance structures and their relationship to ecological
effectriveness \citep{halpern_2013,lpezangarita_2014,mascia_2017-m_} or
benefits to fisheries \citep{krueck_2017-J1}; evaluations of marine
reserves rarely provide a hollistic view of the social-ecological system
\citep{lpezangarita_2014}. Here, we combine causal inference techniques
\citep{depalma_2018} and the social-ecological systems framework
\citep{ostrom_2009-hg} to provide a comprehensive ecological and
socioeconomic evaluation of four community-based marine reserves in
three coastal communities in Mexico.

Marine Reserves in Mexico have been commonly implemented as ``core
zones'' within Biosphere Reserves (BRs) that are administered by the
National Commision of Protected Areas (\emph{Comisión Nacional de Áreas
Marinas Protegidas}, CONANP). While CONANP has made efforts to have a
participatory process, the implementation of these zones is still
characterized by top--down approaches. This motivated Civil Society
Organizations (CSOs) to work with coastal communities to implement
community--based marine reserves \citep{uribe_2010-u2}, which are
usually established within a Territorial Use Rights for Fisheries
(TURFs); thus making them TURF--reserves \citep{afflerbach_2014-HP}.
This bottom--up approach allows fishers to disign their own reserves,
which increases compliance and self--enforcement
\citep{gelcich_2015-Gw,espinosaromero_2014-PY,beger_2004-Y8}. However,
these reserves still lack legal recognition, making them vulnerable to
poaching. In 2014, a new norm \citep{nom} allowed fishers to request the
legal recognition of a community--based reserve under the name of
``Fishing Refugia'' (\emph{Zona de Refugio Pesquero}, FR). This new norm
thus combines bottom--up approaches to design marine reserves, along
with a legal recognition of the management intervention. Since then,
\emph{39} FR have been implemented along the Pacific, Gulfof
Caloifornia, and Mexican Caribbean coastlines, but no formal evaluation
of their effectiveness has taken place.

While there are ecological factors defining the success of a MR
(\emph{i.e.} habitat representation, initial state of protection,
connectivity to other protected areas), their effectivness also depends
on the socioeconomic and governance settings under which they are
implemented. Literature shows that many non-ecological characteristics
can play an equally important role in the effectivenes of MRs. For
example, age of a reserve (\emph{i.e.} time since its implementation),
size, and habitat contained were key to the effectiveness of MRs in
Palau \citep{friedlander_2017-oI}. In the Mediterranean,
\citet{difranco_2016-Xw} identify that surveilance and enfocrement,
presence of a management plan, and involvement of fishers in management
and decision--making along with promotion of sustainable fishing
practices were the key fators that increased stock health and income to
fishers. At a global level, \citet{edgar_2014-UO} indicate that
enforcement, age, size, and isolation were important factors determining
effectiveness of the reserves.

The objective of this work is twofold: i) Provide the first evaluation
of community--based marine reserves in Mexico, and ii) provide a
comprehensive evaluation of the social--ecological system to identify
how sociocoenomic and governance characteristics relate to ecological
effectiveness. With the purpose of providing a hollistic evaluation, we
combine ecological, socioeconomic, and governance indicators. We use
causal inference techniques to provide a measurement of the effect of
the management intervention, and combine it with the social--ecological
systems framework \citep{ostrom_2009-hg}.

@@

We focus our evaluation in three coastal communities from the Pacific
coast of Baja California (n = 1) and the Mesoamerican Reef System (n =
2; Fig \ref{fig:map}). Isla Natividad (IN) lies west of the Baja
California Peninsula (Fig \ref{fig:map}B), where kelp forests
(\emph{Macrocystis pyrifera}) and rocky reefs are the predominant
habitats. The island is home to a fishing cooperative (\emph{Sociedad
Cooperative de Producción Pesquera Buzos y Pescadores de la Baja
California SCL}), that holds a TURF for spiny lobster (\emph{Panulirus
interruptus}). However, other resources like finfish (yellowtail jack,
\emph{Seriola lalandi}), sea cucumber (\emph{Parastichopus
parvimensis}), red sea urchin (\emph{Mesocentrotus franciscanus}), snail
(\emph{Megastraea turbanica} y \emph{M. undosa}), and abalone
(\emph{Haliotis spp}, until 2010) are also important sources of income.
In 2006, the community decided to implement two community--based marine
reserves within their fishing grounds, seeking to recover depleted
stocks of invertebrate species (mainly lobster and abalone). Until
today, these reserves are yet to be legally recognized as Fishing
Refugia.

The other two communities are Maria Elena (ME; Fig \ref{fig:map}C) and
Punta Herrero (PH; Fig \ref{fig:map}D) in the Yucatan Peninsula, where
coral reefs and mangroves are the representative coastal ecosystems. ME
is a fishing camp --visited intermitently during the fishing season--
belonging to the Cozumel fishing cooperative. PH is home to the ``José
María Azcorra'' fishing cooperative. The main source of income to both
communities is the caribbean spinly lobster fishery (\emph{Panulirus
argus}), which is carried out within their respective TURFs. These
communities also target finfish in the off season, mainly snappers
(Lutjanidae) and groupers (Serranidae). ME established eight marine
reserves in 2012, and PH established four marine reserves in 2013. All
these reserves are legally recognized as Fishing Refugia.

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{\label{fig:unnamed-chunk-1}\label{fig:map}Location of the three
coastal communities studied (A). Isla Natividad (B) is located off the
Baja California Peninsula, Maria Elena (C) and Punta Herrero (D) are
located in the yucatan Peninsula.}
\end{figure}

@@

To perform the evaluation of these reserves we use thre sources of
information. Ecological data come from the annual ecological monitorings
of reserve and control areas, carried out by members from each community
and personnel from the Mexican CSO ``Comunidad y Biodiversidad''
(\href{www.cobi.org.mx}{COBI}). These monitorings record richness and
abundances of fish and invertebrate species in the reservs and control
sites. For fish census, size structures are also collected to derive
biomass. We define control sites as regions with habitat characteristics
similar to the corresponding reserves, and that presumably had the same
probability of being selected as reserves during the design phase. From
all the reserves in these three communities, we use the ones that have
data for reserve and control sites before and after the implementation
of the reserve (n = 4) , thus providing us with a
Before--After--Control--Impact (\emph{i.e.} BACI) design that allows us
to capture and control for temporal and spatial dynamics
\citep{depalma_2018,ferraro_2006-oW}. BACI designs and causal inference
techniques have proven effective to evaluate marine reserves, as they
allow us to causaly attribute observed changes to the intervention
\citep{moland_2013-VP,Villasenor-Derbez_2018}.

Socioeconomic data come from landing receipts reported to the National
Commission for Aquaculture and Fisheries (\emph{Comisión Nacional de
Acuacultura y Pesca}; CONAPESCA). Data contain monthly lobster landings
(Kg) and value (MXP) from 2000 to 2014. This information was aggregated
by year, and economic values were adjusted by the Consumer Price Index
\citep{oecd_2017-VV} via Eq \ref{eqn:cpi}.

\begin{equation}
I_t = RI_t\times\frac{CPI_t}{CPI_T}
\label{eqn:cpi}
\end{equation}

Where \(I_t\) represents the adjusted income for year \(t\) as the
product between the reported income for that year and the ratio between
the consumer price index in that year (\(CPI_t\)) to the most recent
year's consumer price index (\(CPI_T\)).

\textbf{Governance data were collected at the community--level, using
key informants to collect the necesary information on X, Y, and Z. Algo
mas de Ere.}

@@

Following a framework that relates reserve objectives to performance
indicators \citep{Villasenor-Derbez_2018}, we use five biological, two
socioeconomic, and five governance indicators to evaluate these marine
reserves Table \ref{table:indicators}.

\begin{table}[H]

\caption{\label{tab:unnamed-chunk-2}\label{table:indicators}List of indicators used to evaluate the effectiveness of marine reserves, grouped by category.}
\centering
\begin{tabular}[t]{l|l}
\hline
Category & Indicador\\
\hline
Biological & Abundance\\
\hline
 & Richness\\
\hline
 & Shannon's diversity index\\
\hline
 & Biomass\\
\hline
 & Abundance of target species (lobsters)\\
\hline
Socioeconomic & Income from target species\\
\hline
 & Landings from target species\\
\hline
Governance & Type of access to the fishery\\
\hline
 & Perceived degree of illegal fishing\\
\hline
 & Reserve surveilance anf enforcement\\
\hline
 & Type of fishing organization\\
\hline
 & Age of the reserve\\
\hline
\end{tabular}
\end{table}

Biological indicators are analyzed with a difference--in--differences
analysis (Eq \ref{eqn:reg_bio}), which allows us to estimate the effect
that the reserve has on the biological indicators by comparing trends
across time and space (\emph{i.e.} reserve and control sites)
\citep{moland_2013-VP,Villasenor-Derbez_2018}. The analysis is performed
with generalized linear models of the form:

\begin{equation}
I = \alpha_{i} + \gamma_{it} Year_t + \beta Zone_i + \lambda_{it} Year_t\times Zone_i + \sigma_jSpp_j + \Pi_{k}Site_k + \epsilon
\label{eqn:reg_bio}
\end{equation}

Where year--fixed effects are represented by \(\gamma_{it} Year_t\),
\(\beta Zone_i\) captures the difference between reserve (\(Zone = 1\))
and control (\(Zone = 0\)) sites, and the interaction term
\(\lambda_{it} Year_t\times Zone_i\) represents the average treatment
effect for each year. When evaluating biomass and abundances, we include
species--fixed effects (\(\sigma_j\)), and when more than one reserve is
present we also include site--fixed effects (\(\Pi_k\)). For abundances
(\emph{i.e.} count data) the model is estimated with a quasipoisson
error distribution.

Socioeconomic indicators are evaluated with a similar approach (Eq
\ref{eqn:soc_reg}), where landings and income before and after the
implementation of the reserve are compared:

\begin{equation}
I = \beta_0 + \beta_1Post
\label{eqn:soc_reg}
\end{equation}

This approach does not allow for a causal attribution of the observed
changes to the reserve, but still allows us to draw important
information that can inform our conclusions. For both approaches, model
coefficients are estimated via ordinary least--squares and
heteroskedastic--robust standard errors \citep{zeileis_2004-7n}.

@@

\textbf{Texto de ere y los SES}
